---
title: "Introduction to grassmapr"
author: "Daniel M Griffith, Rebecca L Powell, Sydney M Firmin, Jen M Cotton, & Christopher J Still"
date: "December 2018"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to grassmapr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction & Objectives
This vignette serves as an introduction to the `grassmapr` R package. The 
workflow that follows also requires learning some basic fuctionality in the 
`raster` R package.

Our goal is to predict plant tissue $\delta$^13^C variation across a landscape. 
The primary driver of variation in plant $\delta$^13^C is the greater isotopic 
fractionation in C~3~ plants compared to C~4~ plants. The physiologically based 
crossover temperature model explains the turnover from C~3~ to C~4~ plants along gradients of temperature (Ehleringer et al. 1997, Collatz et al. 1998, Still et 
al. 2003). 

We will combine the crossover temperature model with data on land cover and 
climate to produce map of C~3~ and C~4~ plants in North America. Then, we will 
apply isotopic endmember values to produce a spatially continuous representation 
of leaf $\delta$^13^C distribution across the land surface. These layers are 
useful for understanding grass biogeography (e.g., Powell et al. 2012, Griffith 
et al. 2015) and for studies seeking to identify the movement of animals (e.g., 
Hobson 1999, Bowen & West 2008).

## Load North America example data

First, load the required R libraries:

```{r, message = FALSE, warning = FALSE, results = "hide"}
library(raster)
library(grassmapr)

```

Next, load the spatial datasets that we will use in this example.
<br>
```{r, message = FALSE, eval = TRUE, warning = FALSE, results = "hide"}
data(temp_NA)     # Mean monthly temperature (deg. C)
data(prec_NA)     # Mean monthly precipitation totals (mm)
data(ndvi_NA)     # Mean monthly NDVI ("greenness")
data(woody_NA)    # Vegetation layer: woody cover (%)
data(cropC3_NA)   # Vegetation layer: C3 crop cover (%)
data(cropC4_NA)   # Vegetation layer: C4 crop cover (%)

```

The rainfall and temperature data are monthly climatologies, so we might want to 
visualize them as annual means. Based on the crossover temperature model, where 
would you expect to find C~4~ grasses?
<br>
```{r, message = FALSE, warning = FALSE, results = "hide", fig.width = 7.5, fig.height = 3.25}

par(mfrow = c(1,2))  
annual_temp <- calc(temp_NA, fun = mean)
plot(annual_temp, 
  main = expression(paste("Mean Annual Temperature (",~degree, "C)", sep = ""))) 

annual_prec <- calc(prec_NA, fun = sum)/100      # rescale to plot
plot(annual_prec, 
  main = expression(paste("Total Annual Precipitation (x10"^2, "mm)", sep = "")))

```
<br> <br>
Before we proceed, we should ensure that all input data have the same geographic 
extent, spatial resolution, and projection.
<br>
```{r, message = FALSE, warning = FALSE, results = "hide"}

compareRaster(temp_NA,
              prec_NA,
              ndvi_NA,
              woody_NA,
              cropC4_NA,
              cropC3_NA)

```

If any of these layers had a different extent or resolution, we could apply the 
function raster::resample() to make the layers compatible with each other. If 
any raster layer had a different projection, the raster::projectRaster() 
function can be applied to transform the layer's projection. 


## Workflow to produce a C~3~ and C~4~ distribution map for North America

First, we will set a C~4~ temperature threshold based on the crossover 
temperature model. Note: In this example we are using mean monthly 
temperature as our climatic variable, but we could set a threshold based on 
other temperature variables, such as maximum monthly temperature. 

Second, we will set a minimum growing season temperature threshold, as well as a 
minimum monthly precipitation threshold. The latter screens for locations that 
have sufficient moisture to support vegetation growth for a given month - as a 
result, locations that lack sufficient precipitation for plant growth 
in a given month are excluded from further analysis - for example, deserts and Mediterranean climates. 
<br>
```{r, message = FALSE, warning = FALSE, results = "hide"}
# Set a C4 temperature threshold based on the COT model (>= 22 deg. C)
C4_temp <- 22
# Set a growing season temperature threshold (>= 5 deg. C)
GS_temp <- 5
# Set a precipitation threshold (>= 25 mm)
GS_prec <- 25
  
```

Next, we create two stacks of climate mask layers to represent (i) months that 
favor C~4~ plants and (ii) months that meet the growing season (GS) criteria. 
Note: These masks classify grid cells that meet climate criteria as 1, and 
grid cells that fail one or more criteria as 0. 
<br>
```{r, message = FALSE, warning = FALSE, results = "hide"}

# Generate monthly C4 climate masks
C4_masks <- mask_climate(temp.stack = temp_NA,
                         temp.threshold = C4_temp,
                         precip.stack = prec_NA,
                         precip.threshold = GS_prec)

# Generate monthly GS climate masks
GS_masks <- mask_climate(temp.stack = temp_NA,
                         temp.threshold = GS_temp,
                         precip.stack = prec_NA,
                         precip.threshold = GS_prec)

```

We can count the monthly masks to generate an annual representation of the total 
number of C~4~ months and total number of growing season months for each 
location. 
<br>
```{r, message = FALSE, warning = FALSE, results = "hide", fig.width = 7.5, fig.height = 3.25}
  
# Count number of months that satisfy each climate criteria
GS_month_total <- count_months(GS_masks)
C4_month_total <- count_months(C4_masks)
  
par(mfrow = c(1,2))
plot(GS_month_total, 
  main = expression("Growing Season (# months)"), zlim = c(0, 12))
plot(C4_month_total, 
  main = expression("C4 Climate (# months)"), zlim = c(0, 12))
  
```
<br> <br>
With these two monthly climate masks - and, optionally, layers of monthly 
"greenness" weights - we can calculate the proportion of the herbaceous layer 
that is C~4~ (i.e., the C~4~ *potential*). Note: This is different than 
predicting *actual* vegetation cover.
<br>
```{r, message = FALSE, warning = FALSE, results = "hide", fig.width = 7.5, fig.height = 3.25}

# Calculate C4 proportion based on C4 climate only
C4_ratio <- calc_C4_ratio(C4_masks, GS_masks)

# Optionally: Calculate C4 proportion based on C4 climate AND vegetation "greenness"
C4_ratio_vi <- calc_C4_ratio(C4_masks, GS_masks, veg.index = ndvi_NA)
  
par(mfrow = c(1,2))
plot(C4_ratio, main = expression(paste("Herbaceous C"[4]~"proportion"))) 


```
<br> <br>
To generate representations of *actual* vegetation cover, we can combine the 
herbaceous C~4~ proportion with other vegetation layers such as woody cover and 
crop cover. Note: The user must provide additional vegetation layers that 
represent *actual* land cover, and these layers must match the geographic 
extent, spatial resolution, and projection of the climate layers.
<br>
```{r, message = FALSE, warning = FALSE, results = "hide", fig.width = 7.5, fig.height = 6}

# Create raster stack of other (non-grass) vegetation layers
veg_layers <- stack(woody_NA, cropC3_NA, cropC4_NA)
  
# Indicate layers that correspond to C4 vegetation
C4_flag <- c(0, 0, 1)

# Indicate layers that correspond to herbaceous vegetation
herb_flag <- c(0, 1, 1)
  
# Calculate plant functional type layers (C4 grass, C3 grass, woody)
pft_cover <- calc_pft_cover(C4.ratio = C4_ratio, 
                            GS.mask = GS_masks, 
                            veg.layers = veg_layers, 
                            C4.flag = C4_flag, 
                            herb.flag = herb_flag)  

par(mfrow = c(2,2))
plot(pft_cover[[1]], 
  main = expression(paste("C"[4]~"Herbaceous cover (%)", sep = "")))
plot(pft_cover[[2]], 
  main = expression(paste("C"[3]~"Herbaceous cover (%)", sep = "")))
plot(pft_cover[[3]], 
  main = expression("Woody cover (%)"))

```


## Turning the vegetation map into an isoscape

Finally, we will generate a vegetation stable carbon ($\delta$^13^C) isoscape by 
applying a simple linear mixing model to each grid cell. From the literature, we 
identify $\delta$^13^C endmember values for each plant functional type layer. 
Isotopic endmembers are weighted by the respective percent vegetation 
composition of each pixel (in this case, C~4~ herbaceous, C~3~ herbaceous, C~3~ 
woody). 

Note: In this example, we have elected to combine the C~4~ and C~3~ crop 
layers with corresponding natural grass layers; however, the functions presented 
here could be easily adapted for custom plant cover types, corresponding to
different isotopic endmember values.


```{r, message = FALSE, warning = FALSE, results = "hide", fig.width = 7.5, fig.height = 3.25}

# d13C endmember vector for PFT layers from the literature
d13C_emb <- c(-12.5, # C4 herb
              -26.7, # C3 herb
              -27.2) # woody

# Apply mixing model to generate d13C isoscape
d13C_iso <- calc_del13C(pft_cover, d13C_emb)

# Standard deviation of d13C endmember means from the literature
d13C_std <- c(1.1, # C4 herb
              2.3, # C3 herb
              2.5) # woody

# Calculate weighted standard deviation of mean d13C values
d13C_iso_std <- calc_del13C(pft_cover, d13C_std)

# Mac Users: substitute "\u2030" for "\211"
par(mfrow = c(1,2))
plot(d13C_iso, main = expression(Plant~{delta}^13*C~"\211"), 
  zlim = c(-27,-12))  
plot(d13C_iso_std, main = expression(StDev~{delta}^13*C~"\211"),
  zlim = c(1.0, 2.4)) 
  
```
<br><br>

## References

Bowen, G. J. and West, J. B. 2008. Isotope landscapes for terrestrial migration 
research. -- In: Hobson, K. A. and Wassenaar, L. I. (eds.), Tracking Animal 
Migration with Stable Isotopes. Academic, pp. 79-105.

Collatz, G. J. et al. 1998. Effects of climate and atmospheric CO~2~ partial 
pressure on the global distribution of C~4~ grasses: present, past, and future. 
-- Oecologia 114: 441-454.

Ehleringer, J. R. et al. 1997. C~4~ photosynthesis, atmospheric CO~2~, and 
climate. -- Oecologia 112: 285-299.

Griffith, D. M. et al. 2015. Biogeographically distinct controls on C~3~ and 
C~4~ grass distributions: merging community and physiological ecology. -- 
Global Ecol. Biogeogr. 24: 304-313.

Hobson, K. A. 1999. Tracing origins and migration of wildlife using stable 
isotopes: a review. -- Oecologia 120: 314-326.

Kohn, M. J. 2010. Carbon isotope compositions of terrestrial C~3~ plants as 
indicators of (paleo) ecology and (paleo) climate. -- Proc. Natl. Acad. Sci. 
107: 19691-19695.

Powell, R. L. et al. 2012. Vegetation and soil carbon-13 isoscapes for South 
America: integrating remote sensing and ecosystem isotope measurements. -- 
Ecosphere 3: 109.

Still, C. J. et al. 2003. Global distribution of C~3~ and C~4~ vegetation: 
carbon cycle implications. -- Global Biogeochem. Cycles 17: 1006.
<br>








